---
title: "Satellite Imagery — DSfE 2025"
author: "Irene Iodice"
date: "May 2025"
output: html_document
---

This script demonstrates how to work with raster (satellite) and vector (shapefile) spatial data in R. It covers:

- Loading and visualizing elevation and nightlight data
- Stacking and aggregating raster images across time
- Comparing changes in light intensity over time
- Cropping and masking rasters using vector boundaries (e.g., Ukraine's border)
- Visualizing and extracting raster data for further analysis


```{r}
# Set working directory
setwd("~/work/Teaching/data-science-for-economists/07-satellite-imagery")

# Load required packages
if (!require("pacman")) install.packages("pacman"); library("pacman")
p_load(terra, sf, stringr, data.table, ggplot2, rnaturalearth)
p_load_current_gh("ropensci/rnaturalearthhires")
```

<!-- # What Is a Raster? Think of it like a spreadsheet laid over the Earth: each cell (pixel) holds a value (e.g., elevation, temperature, light intensity). All cells are evenly spaced. The raster has: resolution (size of each pixel), extent (geographic area it covers), CRS (coordinate reference system) -->


```{r}
# Load and plot an example raster (elevation)
img <- rast(system.file("ex/elev.tif", package = "terra"))
plot(img)
# Which country is this?
```


```{r}
# Load and plot nightlight image (A nightlight .tif file might look like a black-and-white image, but R reads it’s a 2D grid where each cell (pixel) has a numeric value representing light intensity.)
img_nightlight <- rast("input/20210404_4x.tif")
plot(img_nightlight)
# Why you dont see anything? How would you get mote signal?
```

```{r}
# Load all .tif files in folder
list_images <- list.files("input", pattern = "*.tif", full.names = TRUE)
length(list_images)
images <- rast(list_images)

# Filter files for May 2022 and May 2021
list_images_may22 <- list_images[str_detect(list_images, "202205")]
images_may22 <- rast(list_images_may22)

# Calculate median raster across time for each period
images_may22_median <- median(images_may22)
plot(log(images_may22_median))
```
```{r}
images_may21 <- rast(list_images[str_detect(list_images, "202105")])
images_may21_median <- median(images_may21)
plot(log(images_may21_median))
```

```{r}
# Compute pixel-wise difference in light intensity between years
images_may_diff <- images_may21_median - images_may22_median
plot(log(images_may_diff))
```
```{r}
images_may21_mean <- mean(images_may21)
images_may22_mean <- mean(images_may22)


# Extract pixel values for histogram plotting
plot_data_may22 <- terra::values(images_may22_mean, dataframe = TRUE)
plot_data_may21 <- terra::values(images_may21_mean, dataframe = TRUE)
setDT(plot_data_may22)
setDT(plot_data_may21)

# Compare pixel intensity distributions (nightlight proxy)
ggplot() +
  theme_minimal() +
  geom_density(data = plot_data_may22, aes(x = mean), n = 50, color = "red") +
  geom_density(data = plot_data_may21, aes(x = mean), n = 50, color = "blue") +
  scale_y_log10("PDF") +
  scale_x_continuous("Avg. nightlight intensity")
```

<!-- # --- Focused Analysis on Ukraine --- -->

<!-- # In the following steps, we isolate and analyze nightlight intensity data specifically for Ukraine. -->
<!-- # This includes: -->
<!-- # 1. Loading Ukraine’s vector boundary (country shape) from the Natural Earth dataset. -->
<!-- # 2. Ensuring its coordinate reference system (CRS) matches the raster data. -->
<!-- # 3. Cropping the raster to the extent of Ukraine (keeping a rectangular area). -->
<!-- # 4. Masking the raster to the shape of Ukraine (removing everything outside the border). -->
<!-- # This allows us to visualize and quantify changes in light intensity specifically within Ukraine’s borders. -->

 <!-- vector data (country borders) using rnaturalearth and sf packages: -->

```{r}
# Load and plot country borders
plot(ne_countries())
```
```{r}
# Extract Ukraine and convert to sf format
shape_ukraine <- ne_countries(country = "Ukraine")
shape_ukraine <- st_as_sf(shape_ukraine)
plot(shape_ukraine)
```


```{r}

# Check and align coordinate reference systems (CRS)
st_crs(shape_ukraine)
st_crs(images_may22_median)
all.equal(st_crs(shape_ukraine), st_crs(images_may22_median))  # usually returns FALSE
shape_ukraine <- st_transform(shape_ukraine, st_crs(images_may22_median))
all.equal(st_crs(shape_ukraine), st_crs(images_may22_median))  # should now match

```
```{r}
# Crop and mask raster to Ukraine's borders
images_may22_median_ukraine <- crop(images_may22_median, shape_ukraine)
plot(log(images_may22_median_ukraine))
```
```{r}
images_may22_median_ukraine_masked <- mask(images_may22_median, shape_ukraine)
plot(log(images_may22_median_ukraine_masked))
```

