
---
title: "Trade Networks: Metrics and Null Models"
author: "Irene Iodice"
output: html_document
---

## Class Overview

Welcome to today's session: **Trade Networks – Metrics and Null Models**

In this class, we'll explore how to represent and analyze international trade as a **network**, using real trade data from CEPII's BACI dataset. We'll work through the following steps:

1. **Data Preparation** – Load and clean BACI trade data for 2017.
2. **Building the Trade Network** – Construct a directed network where nodes are countries and edges represent trade relationships.
3. **Visualizing the Network** – Examine the global trade structure: identify hubs, assess sparsity, and connectivity.
4. **Network Metrics** – Compute and interpret centrality measures:
   - **Degree centrality**: number of trade partners
   - **Closeness centrality**: distance to other countries in the network
   - **Betweenness centrality**: intermediary or broker roles
5. **Weighted Network Analysis** – Build a weighted network using trade values and reassess centrality.
6. **Null Model Comparison** – Compare the observed network to random graphs (Erdős–Rényi) to understand whether the trade world is "flat". In other words, does trade flow freely and evenly across countries—as it would in a random network—or are there structural barriers and hubs that make global trade more uneven?

### Key Questions
- What makes a country central in global trade?
- How do different metrics reflect economic influence?
- What distinguishes the real trade network from a random one?

You can download the complete data at [this link here](https://www.cepii.fr/CEPII/en/bdd_modele/bdd_modele_item.asp?id=37). 
https://www.cepii.fr/DATA_DOWNLOAD/baci/doc/DescriptionBACI.html

```{r setup, include=FALSE}
required_packages <- c("data.table", "igraph", "ggplot2", "dplyr", "R.utils")

# Install missing packages
new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

# Load all packages
lapply(required_packages, library, character.only = TRUE)

# R Markdown chunk options
knitr::opts_chunk$set(echo = TRUE)
```


## Load and Clean BACI Data

```{r}
data_path <- "~/work/Teaching/data-science-for-economists/05-networks/data/"

file1 <- "BACI_HS07_Y2017_V202001_p1.csv.gz"
file2 <- "BACI_HS07_Y2017_V202001_p2.csv.gz"

conversion <- fread(paste0(data_path, "country_codes_cepii_V2021.csv.gz"))[, .(isocode3 = iso_3digit_alpha, country_code)]

data_baci <- rbind(fread(paste0(data_path, file1)), fread(paste0(data_path, file2)))
data_baci <- data_baci[!grepl("^98", k) & !grepl("^99", k), .(t, i, j, k, v)]

data_baci <- data_baci[conversion, on = .(i = country_code), nomatch = 0][conversion, on = .(j = country_code), nomatch = 0]
setnames(data_baci, old = c("isocode3", "i.isocode3"), new = c("exp", "imp"))

data_baci[, c("i", "j") := NULL]
setnames(data_baci, old = c("t", "k", "v", "exp", "imp"), new = c("year", "product_code", "trade_value", "exp", "imp"))

data_baci <- data_baci[trade_value > 100]

head(data_baci)
```

## 2. Build the Trade Network (Unweighted)

```{r}
# Directed trade graph (no weights)
trade_network <- data_baci %>%
  select(exp, imp) %>%
  distinct() %>%
  graph_from_data_frame(directed = TRUE)
```

**Questions:**
- What kind of relationships do nodes and edges represent in this graph?
- Why should this network be directed?

## 3. Visualize the Network

```{r, fig.width=11, fig.height=11}
plot(trade_network, vertex.size=0.01, vertex.label.cex=0.7, edge.arrow.size=0.1)
```

**Questions:**
- Can you visually identify any hubs?
- Is the network sparse or dense?

## 4. Compute Degree Centrality

```{r}
deg_in <- degree(trade_network, mode = "in")
deg_out <- degree(trade_network, mode = "out")

deg_in_df <- data.frame(country = names(deg_in), degree_in = deg_in)
deg_out_df <- data.frame(country = names(deg_out), degree_out = deg_out)

head(arrange(deg_in_df, -degree_in))  # Top importers
head(arrange(deg_out_df, -degree_out))  # Top exporters
```

**Questions:**
- Who are the top exporters and importers by number of trade partners?
- Do these match your expectations?

## 5. Plot Degree Distribution

```{r}
ggplot(deg_in_df, aes(x = degree_in)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "steelblue", color = "white") +
  labs(title = "IN-Degree Distribution", x = "Out-Degree", y = "Density") +
  theme_minimal()
```

**Questions:**
- What is the shape of the distribution? Is it skewed?
- Are most countries trading with many or few partners?

## 6. Closeness Centrality

```{r}
closeness_in <- data.frame(
  country = names(closeness(trade_network, mode = "in")),
  centrality_in = closeness(trade_network, mode = "in")
)

closeness_out <- data.frame(
  country = names(closeness(trade_network, mode = "out")),
  centrality_out = closeness(trade_network, mode = "out")
)

head(arrange(closeness_in, -centrality_in))
head(arrange(closeness_out, -centrality_out))
```

**Questions:**
- Who are the most "accessible" importers and exporters?
- What does it mean to have high closeness but low degree?

## 7. Betweenness Centrality

```{r}
btw_df <- data.frame(
  country = names(betweenness(trade_network, normalized = TRUE)),
  centrality = betweenness(trade_network, normalized = TRUE)
)

head(arrange(btw_df, -centrality))
```

**Questions:**
- Who plays a broker or intermediary role in the trade network?
- Are there countries with **high betweenness but low degree**?

## 8. Count Nodes, Edges, and Compute Density

```{r}
n_nodes <- gorder(trade_network)
n_edges <- gsize(trade_network)
density <- edge_density(trade_network, loops = FALSE)
# 
n_nodes
n_edges 
# density = (n_edges)/(n_nodes*(n_nodes-1)) - edges here are repeated by all combinations when imp-exp swap order!
density
```

**Questions:**
- What is the density of the trade network?
- What does this tell us about global trade interdependence?


### Create a Weighted Trade Network

```{r}
# Group trade flows by exporter-importer pair and sum their trade values
weighted_edges <- data_baci %>%
  group_by(exp, imp) %>%
  summarise(weight = sum(trade_value, na.rm = TRUE), .groups = "drop") %>%
  na.omit()

# Create a directed graph with edge weights
w_trade_network <- graph_from_data_frame(weighted_edges, directed = TRUE)
```

**Questions:**
- What does assigning **weights** to edges mean in the context of trade?
- How does weighting change the role of countries that have **few but large** trade partners?
- Are “influential” countries by volume the same as those by number of partners?

---

## Weighted Closeness Centrality

```{r}
# Compute weighted in-degree centrality (i.e., total import value)
weighted_in_degree <- data.frame(
  country = names(strength(w_trade_network, mode = "in", weights = E(w_trade_network)$weight)),
  in_strength = strength(w_trade_network, mode = "in", weights = E(w_trade_network)$weight)
) %>%
  arrange(desc(in_strength))

# Show top 6 countries by import strength
head(weighted_in_degree)
```

```{r}
# Weighted closeness OUT: countries that can quickly reach others by trade volume
# Compute weighted out-degree centrality (i.e., total export value)
weighted_out_degree <- data.frame(
  country = names(strength(w_trade_network, mode = "out", weights = E(w_trade_network)$weight)),
  out_strength = strength(w_trade_network, mode = "out", weights = E(w_trade_network)$weight)
) %>%
  arrange(desc(out_strength))

# Show top 6 countries by export strength
head(weighted_out_degree)
```

**Questions:**
- Who are the most central **importers** by weighted degree centrality? Why might they be important?
- Who are the most central **exporters** by weighted degree centrality? How do they compare to the unweighted results?
- Are small, highly connected economies still central when we account for trade volume?
- What kind of country has high **unweighted** closeness but low **weighted** closeness—and vice versa?

**Interpretation:**  
Why might a country have high degree but low closeness in the weighted network?

## Null Model: Erdős-Rényi

```{r}
set.seed(1234)
# Compute observed mean path length
obs_path_length <- mean_distance(trade_network, directed = TRUE)

# Simulate 100 Erdős-Rényi null graphs with same size and density
null_path_lengths <- replicate(100, {
  g_null <- sample_gnp(n = gorder(trade_network), 
                       p = edge_density(trade_network), 
                       directed = TRUE)
  mean_distance(g_null, directed = TRUE)
})

# Compute p-value
p_val <- mean(null_path_lengths <= obs_path_length)

# Plot the null distribution with observed value
null_df <- data.frame(path_length = null_path_lengths)

ggplot(null_df, aes(x = path_length)) +
  geom_histogram(binwidth = 0.001, fill = "lightblue", color = "white") +
  geom_vline(xintercept = obs_path_length, color = "red", linetype = "dashed") +
  annotate("text", x = obs_path_length, y = 5, label = paste("Observed =", round(obs_path_length, 2)), hjust = -0.1, color = "red") +
  ggtitle("Null Model vs Observed Path Length") +
  xlab("Average Shortest Path Length") +
  ylab("Frequency") +
  theme_minimal()

# Display p-value
p_val

```
**Interpretation:**  
What a longer than expected Average Short Path means?

<!-- Trade connections are less efficient than you'd expect by chance. Countries might be geographically distant, or trade might be fragmented due to: Tariffs, trade frictions, or political barriers, Regional clustering (e.g., strong intra-EU trade but weak cross-regional), Supply chains skipping many countries, rather than spanning broadly -->


