---
title: "Event Study: Brexit and Trade"
author: "Irene Iodice"
date: "May 2025"
output: html_document
---

## Learning Objectives
Learning Objectives: By the end of this lab you will be able to:
- Manipulate and summarize dates and time-series objects in R using **lubridate** and **zoo**.
- Visualize trade flows over time and annotate policy events (e.g., Brexit) with **ggplot2**.
- Implement event-study style difference-in-differences and gravity-type regressions with **fixest**.
- Critically evaluate the impact of Brexit on UK exports relative to relevant counterfactuals.

```{r setup, include=FALSE}
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(data.table, stringr, lubridate, countrycode, ggplot2, fixest, etable, zoo)
```

## Laboratory: Playing with Dates in R

In this short exercise, we explore how to handle dates in R using `lubridate` and `zoo`.

```{r}
library(lubridate)
library(zoo)

# lubridate example
d1 <- ymd("2020-01-01")
d2 <- ymd("2020-12-31")

# 1. Create a 60-day window around a date
d <- ymd("2021-01-01")
window <- d + days(-30:30)  # 61-day window centered around d
head(window)
window <- d + months(-2:3) 
head(window)
 ADD HERE THE COMMENT ON HOW TO MAP INTERVAL INTO SMALLER UNITS!
interval(d1, d2)

# zoo example: create a monthly time series
z <- zoo(c(100, 101, 102), order.by = as.Date(c("2020-01-01", "2020-02-01", "2020-03-01")))
plot(z, main = "Simple Zoo Time Series")
```

## Preparing the Data

```{r}

EU27 <- c("DEU", "AUT", "BEL", "DNK", "FIN", "FRA", "GRC", "IRL", "ITA", "LUX", "NLD", "PRT", "ESP", "SWE",
          "MLT", "CYP", "EST", "LTU", "LVA", "CZE", "HUN", "BGR", "ROU", "POL", "SVK", "SVN", "HRV")

files <- list.files("input/monthly_hs2", full.names = TRUE)
data <- fread(cmd = paste0("gunzip -c ", files[1]))
# View(data)
# head(data)
```
```{r}
# Check the data
View(data)
head(data)
data[, uniqueN(Reporter)]
data[, uniqueN(Partner)]
data[, uniqueN(Commodity)]
data[, unique(`Trade Flow`)] 
data[, .N]
```

```{r}
# check trade zeros
data[`Trade Flow Code` == 2, .N]

# check trade zeros
data[, uniqueN(Reporter)] * data[, uniqueN(Partner)] * data[, uniqueN(Commodity)] 
data[`Trade Flow Code` == 2, .N] / (data[, uniqueN(Reporter)] * data[, uniqueN(Partner)] * data[, uniqueN(Commodity)])

```
```{r}
wdir <- getwd()
dir.create(file.path(wdir, "temp"), showWarnings = FALSE)

# read all files
files = list.files("input/monthly_hs2", full.names = T)

# f = files[1]
for (f in files) {
    # print(f)

    # load
    data = fread(f)

    # format and clean data
    data = data[, .(date = Period,
                    origin = Reporter,
                    destination = Partner,
                    hs = `Commodity Code`,
                    flow = `Trade Flow Code`,
                    value = `Trade Value (US$)`)]

    # make dates pretty
    data[, date := ymd(str_c(date, "01"))]
    data[, date := date + months(1) - days(1)]
    # str(data)

    # use country codes
    data[, origin := countryname(origin, "iso3c", warn = FALSE)]
    data[, destination := countryname(destination, "iso3c", warn = FALSE)]
    # countrycode("Germany", "country.name", "iso3c")
    # countrycode("Deutschland", "country.name.de", "iso3c")
    # countryname(c("Deutschland", "Alemania", "Allemagne", "Duitsland"), "iso3c")

    # transform value from integer to numeric
    data[, value := as.numeric(value)]

    # only take obs with non-missing variable
    data = data[complete.cases(data)]

    fwrite(data,
           "temp/monthly_1921.csv.gz",
           compress = "gzip",
           append = T)

}
 
# # alternative
# extract_data = function (...) {}
# map(files, ~ extract_data(.x), .progress = T)

# load full dataset
data = fread("temp/monthly_1921.csv.gz")
data = data[flow == 2]
gc()
```

```{r}
# load full dataset
data = fread("temp/monthly_1921.csv.gz")
data = data[flow == 2]
gc()
data[, .N, by = origin] # greater or equal than 125
data[, .(date, origin)] %>%
  unique() %>%
  .[, .N, by = .(year(date), origin)] %>%
  .[, .(N = sum(N)), by = origin] %>% 
  head()
```
 
```{r}
# brexit impact
data[, unique(hs)]
exports_uk = data[origin == "GBR" & hs == "TOTAL", -c("origin", "hs", "flow")]

check <- data[,
  .(value_total = value[hs == "TOTAL"], 
    value_sum = sum(value[hs != "TOTAL"])),
  by = .(date, origin, destination, flow)
][, match := all.equal(value_total, value_sum)]
```


```{r}
# plot: aggregate over time
plot_data = exports_uk[, .(value = sum(value)), by = date]

plot=ggplot(plot_data) +
    theme_minimal() +
    geom_line(aes(x = date, y = value / 1000000)) +
    geom_vline(aes(xintercept = ymd("2020-01-31")), color = "red") +
    geom_vline(aes(xintercept = ymd("2020-12-31")), color = "red") +
    scale_x_date(name = NULL) +
    scale_y_continuous(name = "Total exports in mn USD",
                       labels = scales::dollar) +
    ggtitle(label = "Total value of UK exports",
            subtitle = "UN Comtrade Data, 2019 – 2021")
ggsave(plot,
       filename = str_c(wdir, "/output/UK_exp.png"),
       width = 20,
       height = 20,
       units = "cm")
plot
```

```{r}
# plot: compare to other countries
plot_data = data[origin %in% c("GBR", "IRL", "ISL", "SWE"), .(value = sum(value)), by = .(date, origin)]
plot_data[, value_norm := value / mean(value[year(date) == "2019"]), by = origin]
```

```{r}
plot_data = data[origin %in% c("GBR", "IRL", "ISL", "SWE") & destination %in% EU27, .(value = sum(value)), by = .(date, origin)]

plot_data[, value_norm := value / mean(value[year(date) == "2019"]), by = origin]

plot_data[, Country := countrycode(origin, "iso3c", "country.name")]
plot = ggplot(plot_data) +
    theme_minimal() +
    geom_line(aes(x = date, y = value_norm, group = Country, color = Country)) +
    geom_vline(aes(xintercept = ymd("2020-01-31")), color = "red") +
    geom_vline(aes(xintercept = ymd("2020-12-31")), color = "red") +
    scale_x_date(name = NULL) +
    scale_y_continuous(name = "Exports to EU compared to average of 2019") +
    ggtitle(label = "Impact of Brexit",
            subtitle = "UN Comtrade Data, 2019 – 2021")

ggsave(plot,
       filename = str_c(wdir, "/output/exp_to_EU27.png"),
       width = 20,
       height = 20,
       units = "cm")
plot
```

## Regression Analysis

```{r}
# regressions instead of eyeball econometrics:all
  # 1. regression export value from the 4 exporters toward EU destinations on Brexit treatment
        # Identification: use va VAriation within country and controlling for (monthly) time trends common to all 4 exporters.
  # 2. the same as 1 but for exports towards all destinations? What do you expect? 
  # 3. the same but including zeros and estimating via PPML
reg_data = data[origin %in% c("GBR", "IRL", "ISL", "SWE") & destination %in% EU27, .(value = sum(value)), by = .(date, origin)]

reg_data[, treatment := (origin == "GBR") * (date > ymd("2020-12-31"))]

reg1 = feols(log(value) ~ treatment | date + origin, data = reg_data)
etable(reg1)

```
```{r}
# all countries
reg_data = data[, .(value = sum(value)), by = .(date, origin)]

reg_data[, treatment := (origin == "GBR") * (date > ymd("2020-12-31"))]

# 162 zeros
reg2 = feols(log(value) ~ treatment | date + origin, data = reg_data)
fixest::etable(reg2)
```
```{r}

reg3 = fepois(value ~ treatment | date + origin, data = reg_data)
fixest::etable(reg3)
```
```{r}
# gravity
reg_data = data[hs == "TOTAL", -c("hs", "flow")]
reg_data[, origin_date := str_c(origin, date)]
reg_data[, destination_date := str_c(destination, date)]
reg_data[, origin_destination := str_c(origin, destination)]

reg_data[, treatment := (origin == "GBR") * (destination %in% EU27)* (date > ymd("2020-12-31"))]

reg4 = feols(log(value) ~ treatment | origin_date + destination_date + origin_destination, data = reg_data)
fixest::etable(reg4)
```

## Exercise

1. Modify the plot to show exports of Ireland only.
2. Try changing the date of treatment and see how the regression coefficients change.
3. Add more EU countries to the comparison.
